---
title: Cumulative improvements in iterated problem solving
author:
  - name: Pierce Edmiston
    email: pedmiston@wisc.edu
    department: Department of Psychology
    university: University of Wisconsin-Madison
  - name: Maxime Derex
    email: maxime.derex@gmail.com
    department: Biosciences Department
    university: University of Exeter
  - name: Gary Lupyan
    email: lupyan@wisc.edu
    department: Department of Psychology
    university: University of Wisconsin-Madison
    last: yes
bibliography: references.bib
csl: apa.csl
abstract: |
  As compared to other animals, humans are particularly skilled at using and
  improving tools and other solutions to problems that were first discovered by
  other people. Although the human capacity for cumulative cultural evolution is
  well-known, the effectiveness of inheritance as a form of problem solving is
  an area in need of further research. We report an experiment designed to
  understand how effectively solutions to problems accumulate over generations
  of problem solving. Using a tool-discovery game, we found that participants
  were consistently able to discover more tools in a 25 minute session than
  their ancestors. Participants who inherited more tools required more time to
  recreate them, but their rate of new tool discovery was not slowed. In
  addition, we show that participants were able to recreate the tools they
  inherited more efficiently than their ancestors, but that inheritance did not
  confer any improvement in future problem solving. We discuss the limitations
  of this work, and motivate future directions.
keywords: "cultural evolution; transmission chain; iterated learning"
output:
  bookdown::pdf_document2:
    template: cogsci.tmpl
    citation_package: natbib
    keep_tex: yes
---
```{r iterated-problem-solving-config, include=FALSE}
library(knitr)
library(crotchet)  # remotes::install_github("pedmiston/crotchet")

opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = "hide",
  fig.width = 3,
  fig.height = 3,
  fig.crop = FALSE,
  fig.pos = "tb"
)

# Read all knitr chunks from external .R file
read_chunk("iterated-problem-solving.R")
```
```{r iterated-problem-solving, include=FALSE}
```

# Introduction {-}

Humans are effective problem solvers, having solved a wide range of problems
related to foraging, hunting, and preparing food, while surviving predators,
each other, and a large range of terrestrial environments [@Boyd:2018diff;
@Fernandez:2001civ]. What has enabled our success in being able to solve such a
diverse set of problems? Some have suggested that the answer lies more in our
ability to inherit knowledge from others than our ability to make discoveries by
ourselves [@Richerson:2005not; @Henrich:2015srt; @Boyd:2018diff]. Humans possess
a number of advanced social learning abilities including teaching through verbal
instruction and imitation that provide reliable ways of transferring problem
solving knowledge across individuals [@Dean:2012gv]. If problem solving
knowledge can be acquired via social learning, then future generations can adapt
and improve it, allowing cultures to accumulate technological complexity over
generations.

However, the ability to learn socially from others is not sufficient to explain
cumulative cultural evolution. Although social learning was once thought to be
rare in the animal kingdom [e.g., @Thorndike:1898], it has now been documented
in a range of species from chimps [@Whiten:1999cic] to fish [@Laland:1997ts]
and even bees [@Alem:2016fu]. If cumulative cultural evolution depended simply
on social learning, we might expect these species to likewise show evidence of
cumulative cultural evolution, yet such evidence is notably lacking
[@Dean:2012gv; @Tennie:2009gaa; but see @Hunt:2003dj; @Sanz:2009ja].

Humans, in contrast, have demonstrated a remarkable ability to adapt and
improve the tools and other innovations discovered by others. The history of
human technology is argued to be better understood as a process of gradual
refinement and repurposing rather than punctuated advances brought by the
discoveries of rare geniuses [@Basalla:1989wq; @Sole:2013cm]. Rapid refinement
of inherited innovations has not always been the case over the course of human
history, as demonstrated by the long periods in the archaeological record of
slow or stagnant growth in stone tool complexity [@Torre:2011jl;
@Lycett:2008di]. Since then, as humans evolved more robust ways of transmitting
cultural information, future generations were able to more quickly learn the
skills honed by their ancestors, thus giving them more time to make
improvements to those technologies [@Sterelny:2012ea; @Kaplan:2002iw]. On this
view, human cultural evolution has been defined not by our ability to copy the
skills of our ancestors, but by the ability to exceed and improve them.

```{r iterated-problem-solving-fig, fig.pos="b", fig.align="center", fig.width=3.5, fig.cap="Iterated problem solving paradigm. Participants were assigned to generations within chains. Each participant completed the same problem solving task for 25 minutes. Participants in generations after the first began the problem solving task with the solutions that were discovered by the previous generation."}
draw_graphviz("iterated-problem-solving", package = "totems")
```

We investigated the human propensity to exceed our ancestors in a problem
solving task using a transmission chain paradigm (Fig.
\@ref(fig:iterated-problem-solving-fig)). Previous research using this paradigm
has found that problem solving performance can accumulate over generations
[@Caldwell:2008gf; @Wasielewski:2014gg; @Zwirner:2015jn]. These experiments
however typically investigate individuals' performance about a single problem
(such as building paper airplanes or baskets). Yet, cumulative cultural
evolution allows humans to solve more diverse and ever larger sets of problems.
Larger sets of problems can only be solved when larger amounts of information
are transmitted between generations, which is likely to result in increasing
acquisition costs for learners [@Mesoudi:2011fn]. Despite its pivotal importance,
the effects of increasing amount of information on learners' performance have
not been investigated experimentally.

Here we allow future generations to inherit symbolic information
("recipes") about how to recreate the tools that had been discovered by an
ancestor and measure the ability of participants to recreate and exceed
the tools they inherited. In addition to asking whether participants were able
to exceed the total number of tools discovered by their ancestor, we also asked
whether inheritance influenced the way in which future problems were solved. To
answer this question, we analyzed whether participants who inherited more tools
from their ancestors were more or less effective at discovering new tools. We
also analyzed the guessing strategies used by participants who benefited from
inheritance as compared to first generation participants who did not inherit
from any ancestor. These analyses are used to address potential downstream
consequences to iteratively inheriting from a previous generation.

# Methods {-}

To understand how solutions to problems accumulate through vertical
transmission, we used a transmission chain paradigm where participants were
assigned to a single generation within a four-generation chain. Each
participant attempted the same tool discovery task for 25 minutes. The recipes
for how to create the tools that each participant had discovered by the end of
the session were passed on to be inherited by a participant in the next
generation of the chain. Thus, participants assigned to generations after the
first began the experiment with information about how to create the tools
inherited from the previous generation.

Participants played the "Totem" game adapted from @Derex:2015cbb. Their task
was to discover how to build tools with the ultimate goal of creating "a sacred
totem to appease the gods." To build a totem, participants first needed to
construct an axe out of three independently discovered tools: a refined stick
used as a handle, a sharpened rock for the blade, and a string wound from bark
fibers for binding (Fig. \@ref(fig:landscape-sample-1)). More advanced tools
produce larger and more intricate totems.

```{r landscape-sample-1, fig.height=6, fig.cap="A sample of the solution landscape. The top row of 6 items were available to participants at the start of the game. Tools could be produced through the combination of different items (more than one arrow points to the item) or the combination of the same items (a single arrow points to the item). The axe is required to construct the first totem pole."}
draw_image("landscape-sample.gv", package = "totems")
```

Participants discovered new tools by combining existing items. Participants
could refine individual items, or combine up to four items at a time (with
replacement), meaning the initial six items could form a total of
`r methods$n_unique_guesses_6` combinations. Of all possible combinations, very
few resulted in new items. For example, of all the guesses that could be formed
from the initial items, only three (`r methods$n_unique_guesses_6_pct`%)
yielded new tools.

As tools are accumulated, the number of possible combinations that can be made
with those tools increases exponentially such that the discovery of later tools
was less likely to happen by chance alone. Based on previous research using
this task, we know that participants are far more likely to make some guesses
than others, indicating they are using common knowledge acquired outside the
lab to generate combinations. For example, once discovering an axe,
participants quickly discover that they can use the axe to chop down a tree,
regardless of the other tools they may have at the time. At the same time,
participants do not find all tools equally intuitive, and the difference in
combinatorial complexity should not be ignored. In our results, we report
performance based on both measures.

Once a tool was discovered, the recipe for its production---a list of the items
that had to be combined in order to produce the tool---was recorded in an
innovation record. Participants could review their past innovations and see the
recipes for their previous discoveries. Participants assigned to generations
after the first inherited the innovation record of the previous generation
participant. From the beginning of the experiment, these participants could
review the recipes for all the innovations that had been discovered by their
ancestor. Note that the participants inherited the recipes, but not the tools
themselves. In order use these tools in further combinations, the tools and all
of their constituent parts first had to be recreated.

## Participants {-}

Participants were recruited from the UW-Madison student body and received
course credit in exchange for participation. Each participant was assigned to
a single generation of a four-generation chain. Data was collected
for a total of `r exp1$n_teams` complete chains (N=`r exp1$n_participants` participants).

# Results {-}

Our results are presented in three sections. First we report the total number
of tools discovered by each generation along chains. Second we report the number
of new tools discovered relative to the number of tools inherited (as opposed
to generation). Also in the second section, we quantify the amount of time each
participant spent recreating inherited tools versus discovering new ones, and
test whether the number of tools inherited had an impact on the rate of new
tool discovery. In the last section, we compare first generation participants
who did not inherit from anyone to participants in generations 2-4 who inherited
at least some tools from an ancestor in an attempt to understand whether inheritance
confers any benefit to problem solving beyond the inherited solutions.

## Tools by generation {-}

We found that participants in later generations were able to discover more
tools in the same amount of time than their ancestors (Fig.
\@ref(fig:innovations-by-generation)). To quantify these gains, we fit
a hierarchical regression model to the number of tools discovered in each
generation with polynomial contrasts for generation and random effects for
chain. On average, second generation participants were able to discover
`r exp1$gen0_slope` more tools than first generation participants,
`r exp1$gen0_slope_stats`. This effect decreased by `r exp1$gen0sqr_slope` each
generation for third and fourth generation participants,
`r exp1$gen0sqr_slope_stats`.

(ref:innovations-by-generation) Tools by generation. Each of the thin blue lines is a chain. The thick black line shows the model predictions with \pm1 standard error.

```{r innovations-by-generation, fig.cap="(ref:innovations-by-generation)"}
innovations_by_generation_plot
```

It is worth noting that as tools are accumulated, the number of possible
combinations that can be produced increases exponentially. As a consequence,
the discovery of later tools was less likely to happen by chance alone. To take
the size of the combinatorial space into account, rather than scoring each tool
equally, we instead scored each tool based on the size of the combinatorial
space at the time it was discovered. Refitting the same hierarchical regression
model as above, this time predicting the sum of tool scores discovered in each
generation, we again found that tool scores increased linearly with each
generation, `r exp1$difficulty_by_generation_mod_slope_stats`, but in this
model, the improvement in tool score was not found to decrease for later
generations, `r exp1$difficulty_by_generation_quad_mod_slope_stats`.

## Tools by inheritance size {-}

(ref:inheritance-size) Tools by inheritance size. A. Number of tools created relative to those inherited. The dotted line is a reference with slope=1 such that points above the line indicate future generations exceeding their ancestors. B. Number of new tools relative to those inherited. The same reference line is now shown horizontally. The error range shows the model predictions with \pm1 standard error.

```{r inheritance-size, fig.height=6, fig.cap="(ref:inheritance-size)"}
grid.arrange(
  innovations_created_and_inherited_plot + ggtitle("A"),
  new_innovations_plot + ggtitle("B"),
  ncol = 1
)
```

Because there is no difference between a second generation participant who
inherits 10 tools and a fourth generation participant who inherits the same
tools, we also looked at problem solving performance relative to the number
of tools that were inherited regardless of generation (Fig.
\@ref(fig:inheritance-size)). As the number of inherited tools increased, the
number of new tools discovered decreased, `r exp1$inheritance_size_slope_stats`
(Fig. \@ref(fig:inheritance-size)B).

Participants who inherited more tools also required more time to recreate those
tools. Participants took on average `r exp1$mean_learning_time_min` minutes of the 25
minute session (`r exp1$proportion_learning_time`%) to recreate the inherited
tools---a portion of the experiment we refer to as the learning period. The
length of the learning period correlated positively with the number of inherited
tools, _r_ = `r exp1$learning_cor` (Fig. \@ref(fig:learning-rates)).

```{r learning-rates, fig.cap="Learning rates. Correlation between the number of tools inherited and the time it took to recreate the inherited items. Outliers who were appear unwilling or unable to recreate the inherited items are shown as X's, but included in all analyses."}
learning_rates_plot
```

We next asked whether inheriting more tools had an impact on the rate of new
tool discovery, controlling for the length of time spent recreating inherited tools.
We fit a hierarchical regression model predicting the number of new tools
discovered relative to the amount of time out of the 25 minute session
available to discover new tools (Fig. \@ref(fig:playing-time)).
The overall discovery rate was `r exp1$minutes_per_new_innovation` minutes per
tool (`r exp1$new_innovations_per_minute` innovations per minute),
`r exp1$playing_time_slope_stats`. This rate was not found to vary based on
the number of inherited tools, as revealed by comparing a model predicting
novel tools from discovery time alone to one predicting novel tools
from the interaction between discovery time and inheritance size,
`r exp1$playing_time_modcomp`.

(ref:playing-time) New tool discovery rates. Discovery time is the amount of time out of a 25 minute session dedicated to discovering new innovations that were not discovered by an ancestor. The line shows the predictions of the hierarchical regression model with \pm1 standard error. The slope of this line did not significantly vary based on the number of inherited tools. Participants marked with X's are the same as in Fig. \@ref(fig:learning-rates).

```{r playing-time, fig.cap="(ref:playing-time)"}
playing_time_plot
```

## Guesses per tool {-}

In this section, we compared first generation participants who did not inherit
from anyone to participants in generations 2-4 who inherited at least some
tools from an ancestor. We compared these two groups in terms of the number of
guesses required to discover each tool.

To count the number of guesses that were required for each tool, we tallied all
guesses made from the moment in which a new tool was eligible for discovery
until that tool was discovered. A new tool was eligible for discovery once all
of the items required to produce the new tool had been discovered.

We fit a hierarchical linear model predicting the number of guesses per tool
based on generation (Generation 1, Generations 2-4) with random effects by
tool. Tools that have been produced by Generations 2-4 were assigned to one of
two classes: those that were inherited from an ancestor, and those that were
discovered through a trial-and-error process. This allowed us to test whether
the benefit to inheritance was restricted to reducing the number of guesses for
inherited items, or whether inheritance had any effect on future guessing
behaviors.

As expected, participants from generations 2-4 made fewer guesses per tool than
when the same tools were attempted in the first generation,
`r exp1$guesses_per_item_by_inheritance` (Fig. \@ref(fig:guesses-per-innovation-1)A). This effect demonstrates the benefit of
inheriting from a previous generation in providing a shortcut to discovering
these tools.

However, we did not find any evidence that inheritance had an effect on the
number of guesses per new tool, `r exp1$guesses_per_new_item_by_inheritance_mod` (Fig.
\@ref(fig:guesses-per-innovation-1)B). This finding suggests that although
inheritance benefits participants in recreating inherited solutions, it does
not confer any benefits to future problem solving.

(ref:guesses-per-innovation-1) Guesses per tool by participant generation. Each line is the average number of guesses it took to discover a particular tool. Error bars show \pm1 standard error of the model predictions. A. Inherited tools. B. Discovered tools, not inherited from an ancestor.

```{r guesses-per-innovation-1, fig.height=6, fig.cap="(ref:guesses-per-innovation-1)"}
grid.arrange(
  guesses_per_item_by_inheritance_plot + ggtitle("A"),
  guesses_per_new_item_by_inheritance_plot + ggtitle("B"),
  ncol = 1
)
```

# Discussion {-}

We found that participants were consistently able to solve more problems in
a single 25 minute session than their ancestors, and thus were able to
cumulatively improve upon the solutions they inherited. All participants were
expected to be able to recreate the tools they inherited, but whether they
could discover new tools, beyond those inherited, was unknown. Given the
combinatorial complexity of the solution landscape, participants were unlikely
to strike upon beneficial combinations by guessing at random. Because of this,
some participants were unable to discover any new tools. But most did discover
new tools, even when inheriting an already large number of previously
discovered tools.

We also explored the impact of inheriting solutions on problem solving
performance. We found that participants who inherited more tools tended to
discover fewer new tools than their ancestors, suggesting that later generation
participants had a harder time exceeding their ancestors. Part of the reason is
that later generation participants needed more time to recreate the tools they
inherited. Controlling for the amount of time each participant had to discover
new tools (as opposed to recreating inherited tools) did not reveal an effect
of inheritance size on the rate of new tool discovery. Finally, we investigated
whether the benefit of inheritance in terms of guesses per tool extended beyond
the inherited tools, and found that although inheritance clearly reduced the
number of guesses required for inherited tools, it did not confer any benefit
to future problem solving performance.

Our conclusions are limited by the design of the solution landscape in the
Totem game, and the restriction in our methods to a single problem solving
strategy. The sparsity of the solution landscape, where many combinations can
be made but very few yield new tools, indicates that in order to succeed
participants must use prior knowledge to help form combinations that are most
likely to yield new tools. This challenges the notion that the difficulty of a
particular tool is directly related to its combinatorial complexity. In
addition, we believe the accumulation of problem solving knowledge over
generations must be compared with the accumulation of problem solving knowledge
through other forms of problem solving that do not involve vertical
transmission.

More than any other animal, humans are particularly skilled at inheriting and
improving tools and other solutions to problems, but whether the ability
to inherit from others has effects on problem solving beyond giving a head
start to individual learning is not known. Although much work is still
needed to fully understand the human propensity for cumulative cultural
evolution, we believe our research is a valuable contribution to ongoing
efforts to understand how and why human culture is so integrally cumulative.

# Open science practices {-}

The materials, data, and code used to support the conclusions of this paper are
available via the Open Science Framework page for this research
at [osf.io/vf2wk](https://osf.io/vf2wk), DOI: 10.17605/OSF.IO/VF2WK.
