---
title: Cumulative improvements in iterated problem solving
author: 
  \textbf{Pierce Edmiston}, Maxime Derex, Gary Lupyan
bibliography: references.bib
csl: apa.csl
output:
  beamer_presentation:
    theme: metropolis
    template: slides.tmpl
nocite: |
  @Boyd:2011gp
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(crotchet)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
read_chunk("iterated-problem-solving.R")
```
```{r iterated-problem-solving, include=FALSE}
```
```{r config, include=FALSE, cache=FALSE}
t_$base_theme$text$size <- 20
theme_set(t_$base_theme)
```

# The cultural inheritance of problem solving knowledge

Why are humans effective problem solvers?

- We evolved specialized intelligences [@Pinker:2010gw].
- **We developed systems of cultural inheritance** (Boyd et al., 2011).

<!--
It seems relatively uncontroverial to say that humans are effective problem
solvers, but it is worth considering what the evidence actually is, that we are
good at problem solving. One way to think about it, from an evolutionary
perspective, is that our species, and no other, has been able to solve the
problems necessary for survival in nearly every terrestrial environment on the
planet. No other single species comes close. So what are the explanations for
why humans are so good at problem solving?

One answer to this question is that we evolved specialized intelligences.
On this view, we are effective problem solvers because our brains evolved
so that individuals could be smart and solve problems on their own.

An alternative view is that we owe our success in problem solving in large
part to our systems of cultural inheritance. On this view, it's not our
ability to solve problems as individuals that is all that impressive.
Instead, it's our ability to copy other people that makes us adaptive
problem solvers. Humans may be smart, but cultures are smarter.
-->

# Lost European explorer experiments[^lostexplorer]

> Starvation on nardoo is by no means unpleasant, but for the weakness one
  feels, and the utter inability to move oneself, for as the appetite is
  concerned, it gives me the greatest satisfaction.

```{r nardoo, out.height="40%", fig.cap="The nardoo acquatic fern, native to Australia."}
crotchet::draw_image("img/nardoo.jpg")
```

[^lostexplorer]: [@Boyd:2018diff; @Boyd:2011gp; @Henrich:2015srt]

<!--
A favorite source of evidence for why cultures are more adaptive than
individuals comes from what Rob Boyd has termed the Lost European explorer
experiments. This is where a group of intelligent and well-equipped
adventurers somehow gets stranded in a foreign land, and the test is whether
they can survive off the land long enough to make it back home. The irony of
these experiments is that the explorers often die right alongside indigenous
populations who have somehow figured out how to thrive in the same
environment.

An example of this is the Burke and Wills expedition across Australia. As they
ventured across inland Australia they were running out of supplies, but they
ran into a group of Aborigines that was surviving off the land just fine. They
saw that these Aborigines survived largely by eating an acquatic fern called
nardoo, pictured here. The Aborigines prepared this plant by mashing it into
a paste and forming it into little cakes. But all the explorers saw was that
the Aborigines were eating a lot of nardoo, and so they assumed it wasn't
harmful. It turns out that the food preparation process removed toxins from
the nardoo, and when eaten unprepared, it caused a form of thamine deficiency
that ultimately lead to their deaths.

Here is a passage from Wills diary shortly before he died, describing how
odd it was to die of starvation with a full belly.

> Starvation on nardoo is by no means unpleasant, but for the weakness one
  feels, and the utter inability to move oneself, for as the appetite is
  concerned, it gives me the greatest satisfaction.

The Lost European explorer experiments demonstrate that cultures seem to be
better at solving the problems necessary for survival than any individual
plopped down in the same environmental conditions.
-->

# The adaptiveness of cultural inheritance[^br]

Cultural inheritance is adaptive because **even smart people die.**  
Cultural inheritance enables problem solving to extend far longer than any lifetime.  

```{r adaptiveness, engine="dot"}
digraph {
    rankdir = LR
    node[shape=circle style=filled label=""]
    bgcolor="transparent"
    
    d1, i2  // set row order last to first
    
    d1 -> d2 -> d3 -> d4 -> d5
    i1 -> i2
    d5, i2[style=invis]

    i1[fillcolor="#66C2A5"]

    d1[fillcolor="#7483A7"];
    d2[fillcolor="#8DA0CB"];
    d3[fillcolor="#A1B1D4"];
    d4[fillcolor="#B6C2DD"];
}
```

[^br]: [@Boyd:1985cu; @Richerson:2005not]

<!--
Why is cultural inheritance adaptive? Why is it more beneficial to be able to
copy someone else than it is to be able to solve problems as individuals?

One of the basic arguments given for why cultural inheritance is adaptive is
that even smart people die.

If you cannot copy, you can only accomplish so much on your own in a single
lifetime, no matter how smart you are. If you can copy, you can instead spend
your life improving what other people invented.

What this means is that problem solving can extend far longer than any single
lifetime. Eventually, cultures can solve problems that individuals cannot.
-->

# How likely are future generations to exceed their ancestors?

- Does cultural inheritance accumulate additively?

# The Totem Game [@Derex:2015cbb]

```{r gameplay}
crotchet::draw_image("gameplay", package = "totems")
```

# Solution landscape

```{r landscape-sample, out.height="90%"}
draw_image("landscape-sample.gv", package = "totems")
```

# Tools by generation

```{r fig3, fig.width=5, fig.height=5, out.height="90%"}
innovations_by_generation_plot
```

# Tools by inheritance size

```{r fig4}
grid.arrange(
  innovations_created_and_inherited_plot + ggtitle("A"),
  new_innovations_plot + ggtitle("B"),
  ncol = 2
)
```

# Learning times

```{r fig5, fig.width=5, fig.height=5, out.height="90%"}
learning_rates_plot
```

# Discovery rates

```{r fig6, fig.width=5, fig.height=5, out.height="90%"}
playing_time_plot
```

# Guesses per tool

```{r fig7}
grid.arrange(
  guesses_per_item_by_inheritance_plot + ggtitle("A"),
  guesses_per_new_item_by_inheritance_plot + ggtitle("B"),
  ncol = 2
)
```

# Diachronic collaboration experiment summary

1. How likely were participants able to exceed their ancestors?
    - Participants were able to solve more problems in 25 minutes than their ancestors.
2. Does inheritance have an impact on new tool discovery?
    - No effect on the rate of new tool discovery.
    - No effect on the number of guesses required for new tools.

**Limitation:** How does diachronic collaboration compare to alternative strategies?


# Summary: Totem game

1. Diachronic collaborators are able to exceed their ancestors.
2. Simple diachronic collaboration is comparable to alternatives.
3. Diachronic collaboration may reduce the cost of fixation.
4. Diachronic collaboration may be more scalable than synchronic
   collaboration.

# References {.allowframebreaks}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.25in}
\noindent
\footnotesize
