---
title: Cumulative improvements in iterated problem solving
author: Pierce Edmiston, Maxime Derex, Gary Lupyan
bibliography: references.bib
csl: apa.csl
output:
  beamer_presentation:
    theme: metropolis
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(crotchet)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
theme_set(theme_minimal(base_size = 20))
```

# The cultural inheritance of problem solving knowledge

Why are humans effective problem solvers?

- We evolved specialized intelligences [@Pinker:2010gw].
- We developed systems of cultural inheritance [@Boyd:2011gp].

# Lost European explorer experiments[^lostexplorer]

```{r nardoo, out.height="40%", fig.cap="The nardoo acquatic fern, native to Australia."}
crotchet::draw_image("img/nardoo.jpg")
```

> [S]tarvation on nardoo is by no means unpleasant, but for the weakness one
  feels, and the utter inability to move oneself, for as the appetite is
  concerned, it gives me the greatest satisfaction.

[^lostexplorer]: [@Boyd:2018diff; @Boyd:2011gp; @Henrich:2015srt]

# The adaptiveness of cultural inheritance[^br]

Cultural inheritance is adaptive because **even smart people die.**  
Cultural inheritance enables problem solving to extend far longer than any lifetime.  

[^br]: [@Boyd:1985cu; @Richerson:2005not]

# References {.allowframebreaks}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.25in}
\noindent
\footnotesize
