---
title: Cumulative improvements in iterated problem solving
author: 
  \textbf{Pierce Edmiston}, Maxime Derex, Gary Lupyan
bibliography: references.bib
csl: apa.csl
output:
  beamer_presentation:
    theme: metropolis
    template: slides.tmpl
nocite: |
  @Boyd:2011gp
---
```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(crotchet)
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
read_chunk("iterated-problem-solving.R")
```
```{r iterated-problem-solving, include=FALSE}
```
```{r config, include=FALSE, cache=FALSE}
t_$base_theme$text$size <- 20
theme_set(t_$base_theme)
```

# The cultural inheritance of problem solving knowledge

Why are humans effective problem solvers?

- We evolved specialized intelligences [@Pinker:2010gw].
- We developed systems of cultural inheritance (Boyd et al., 2011).

```{r human-environments, fig.height=5, fig.cap="Human settlements in the Moroccan desert, the Brazilian rainforest, and the Canadian arctic."}
gridExtra::grid.arrange(
  crotchet::read_image("img/moroccan-desert.jpg", width = 0.9),
  crotchet::read_image("img/amazon-rainforest.jpg", width = 0.9),
  crotchet::read_image("img/inuit-arctic.jpg", width = 0.9),
  nrow = 1
)
```

<!--
It seems relatively uncontroverial to say that humans are good problem
solvers, but it is worth considering what the evidence actually is, that we are
good at problem solving. One way to think about it, from an evolutionary
perspective, is that our species, and no other, has been able to solve the
problems necessary for survival in nearly every terrestrial environment on the
planet.

Here are just a few examples: human settlements in the Moroccan desert, the
Brazilian rainforest, and the Canadian arctic. No other single species is able
to survive in such diverse environments. So what are the explanations for why
humans are so good at solving different kinds of problems?

One answer to this question is that we evolved specialized intelligences. On
this view, we are effective problem solvers because our brains evolved under
selection pressures to enable individuals to be smart and solve different kinds
of problems on their own.

An alternative view is that we owe our success in problem solving in large part
to our systems of cultural inheritance. On this view, it's not our ability to
solve problems as individuals that is all that impressive. Instead, it's our
ability to copy other people and benefit from their work that makes us adaptive
problem solvers. Humans may be smart, but cultures are smarter.
-->

# Lost European explorer experiments[^lostexplorer]

> "Starvation on nardoo is by no means unpleasant, but for the weakness one
  feels, and the utter inability to move oneself, for as the appetite is
  concerned, it gives me the greatest satisfaction."

```{r nardoo, out.height="40%", fig.cap="The nardoo acquatic fern, native to Australia."}
gridExtra::grid.arrange(crotchet::read_image("img/nardoo-leaves.jpg", width = 0.9),
                        crotchet::read_image("img/nardoo-seeds.jpg", width = 0.9),
                        nrow = 1,
                        widths = c(0.467, 0.533))
```

[^lostexplorer]: [@Boyd:2018diff; @Boyd:2011gp; @Henrich:2015srt]

<!--
A favorite source of evidence for why cultures are more adaptive than
individuals comes from what Rob Boyd has termed the Lost European explorer
experiments. This is where a group of intelligent and well-equipped
adventurers somehow gets stranded in a foreign land, and the test is whether
they can survive off the land long enough to make it back home. The irony of
these experiments is that the explorers often die right alongside indigenous
populations who have somehow figured out how to thrive in the same
environment.

An example of this is the Burke and Wills expedition across Australia. As they
ventured across inland Australia they were running out of supplies, but they
ran into a group of Aborigines that was surviving off the land just fine. They
saw that these Aborigines survived largely by eating an acquatic fern called
nardoo, pictured here. The Aborigines prepared this plant by mashing it into
a paste and forming it into little cakes. But all the explorers saw was that
the Aborigines were eating a lot of nardoo, and so they assumed it wasn't
harmful. It turns out that the food preparation process removed toxins from
the nardoo, and when eaten unprepared, it caused a form of thamine deficiency
that ultimately lead to their deaths.

Here is a passage from Wills diary shortly before he died, describing how
odd it was to die of starvation with a full belly.

> Starvation on nardoo is by no means unpleasant, but for the weakness one
  feels, and the utter inability to move oneself, for as the appetite is
  concerned, it gives me the greatest satisfaction.

The Lost European explorer experiments demonstrate that in some cases cultures
seem to be better at solving problems than any individual plopped down in the
same environmental conditions.
-->

# The adaptiveness of cultural inheritance[^br]

Cultural inheritance is adaptive because **even smart people die.**  
Cultural inheritance enables problem solving to extend far longer than any lifetime.  
_How effectively does problem solving accumulate over generations?_

```{r adaptiveness, engine="dot"}
digraph {
    rankdir = LR
    node[shape=circle style=filled label=""]
    bgcolor="transparent"
    
    d1, i2  // set row order last to first
    
    d1 -> d2 -> d3 -> d4 -> d5
    i1 -> i2
    d5, i2[style=invis]

    i1[fillcolor="#66C2A5"]

    d1[fillcolor="#7483A7"];
    d2[fillcolor="#8DA0CB"];
    d3[fillcolor="#A1B1D4"];
    d4[fillcolor="#B6C2DD"];
}
```

[^br]: [@Boyd:1985cu; @Richerson:2005not]

<!--
Why is cultural inheritance adaptive? Why is it more beneficial to be able to
copy someone else than it is to be able to solve problems as individuals?

One of the basic arguments given for why cultural inheritance is adaptive is
that even smart people die. If you cannot copy, you can only accomplish so much
on your own in a single lifetime, no matter how smart you are. If you can copy,
you can instead spend your life improving what other people invented. What this
means is that problem solving can extend far longer than any single lifetime.
Eventually, cultures can solve problems that individuals cannot.

This argument makes sense to me, but it also assumes something that I think
should not be ignored, and that is that, on this view, problem solving is
relatively unaffected by cultural inheritance. A person in generation 1 has the
same problem solving capacity as a person in generation 4. All cultural
inheritance gives you is a head start, but it has no effect on future problem
solving.

What I want to do today is convince you that we should not just assume that 
problem solving adds up over generations, but that we should test how effectively
problem solving performance improves over generations of repetition.
-->

# Increasing acquisition costs[^var]

```{r mesoudi-2011-fig2a}
crotchet::draw_image("img/mesoudi-2011-fig2a.png")
```

[^var]: [Fig. 2A, @Mesoudi:2011fn]

<!--
One reason why cultural inheritance may not just add up over generations
is related to the idea of increasing acquisition costs. This is the idea that
before you can solve any new problems, first you have to re-learn what everyone
before you has already done. As the amount you need to re-learn increases,
you have less time to discover new things.

Here's a figure from a paper by Alex Mesoudi showing the relationship between
when a mathematical concept was originally discovered and the age at which that
concept is conventionally learned. Note that the x-axis is logorithmic. The
basic idea is that masters students in mathematics, in their early 20s, are
still learning ideas that may be over 100 years old, in part because they
need to learn the foundations of mathematics before they can learn the most
recent innovations.

Whether this sort of historical recapitulation is necessary for educational
practice is debatable, but what I want you to take away from this example is
that in some cases, before any new progress can be made, new generations first
have to learn what has been done previously, and this might take a long time.
-->

# How does problem solving performance accumulate?

```{r cumulative-culture, engine="dot"}
digraph {
    rankdir = LR
    node[shape=circle style=filled label=""]
    bgcolor="transparent"
    
    d1 -> d2 -> d3 -> d4 -> d5
    d5[style=invis]

    d1[fillcolor="#7483A7"];
    d2[fillcolor="#8DA0CB"];
    d3[fillcolor="#A1B1D4"];
    d4[fillcolor="#B6C2DD"];
}
```

Experimental models of cumulative culture:

- building paper airplanes
- knapping stone tools
- constructing baskets
- solving puzzles on a computer

<!--
The purpose of my research was to better understand this question: how
does problem solving accumulate over generations?

The idea that each new generation is able to build upon the accomplishments
of the previous one is generally known as "cumulative culture". Cumulative
culture seems obvious looking around today. Technologies like cars, computers,
and spaceships are only possible because of generations of accumulated
innovations. But it's only recently that we have really tried to establish
good experimental models of cumulative culture. While we all may agree that
cumulative culture occurs at the longest timescales, it's still an open
question whether we learn anything about human problem solving by studying
cumulative culture at the shortest timescales.

Previous experiments investigating cumulative culture have used a number of
different experimental models. For example, if you watch someone build a paper
airplane and fly it, and then you attempt the same task, chances are you can
build a paper airplane that will fly a bit further than your ancestor, at least
through a handful of generations.

This sort of accumulation depends heavily on the task, of course. People
have also done experiments where participants learn how to knap stone tools,
which is a much more complicated task than building a paper airplane.
For these experiments, participants don't improve upon the stone tool design,
but we are able to better understand how reliably skills are passed on
to the next generation.

Most recently, experimental models of cumulative culture have started to
look more like cognitive psychology experiments in that they have people
solve puzzles on a computer where aspects of the problem solving task can
be more tightly controlled than in an experiment involving the creation
of physical artifacts.
-->

# The Totem Game [@Derex:2015cbb]

```{r gameplay, out.height="90%"}
draw_image("gameplay", package = "totems")
```

<!--
An example of this is the Totem game. In the Totem game, participants attempt
to build a totem pole first by combining these resources to make tools, and using
those tools to create other tools, and so on and so on, until they can chop
down a tree, and carve and paint a totem pole.

In this game, most of the combinations participants try do not yield new tools,
and so the test is how effectively participants can search the space of possible
combinations without making the same incorrect guesses over and over again.
-->

# Solution landscape

```{r landscape, out.height="80%"}
gridExtra::grid.arrange(
  crotchet::read_image("landscape-sample.gv", package = "totems"),
  crotchet::read_image("recipes.gv", package = "totems"),
  nrow = 1
)
```

<!--
What makes this game interesting is that as tools are accumulated, the space
of possible combinations increases exponentially.

Here I'm showing the first 6 generations of tools, leading to the first axe,
which is required to make the simplest kind of totem pole. The landscape goes
on beyond this point, but in reality, in just 25 minutes, participants have
a hard time getting even this far.

On the right, I'm showing the recipes for the tools: the combinations of items
participants had to make in order to discover the tool. So first they had to
combine two stones to make a more refined rock, and a stone with an antler
to make a club, and so one. Some items could be refined all by themselves,
so you could break a branch off a tree, and refine a branch into a stick by
trying these items individually. Most tools involved the combination of between
2 and 3 items.

Presumably, with more items, the game gets harder and harder, so we can ask
how individuals who inherit a bunch of items from an ancestor are able to
exceed them. Are they able to overcome the acquisition cost of recreating
the previous tools?
-->

# Methods

```{r cumulative-culture, engine="dot"}
```

- 42 complete chains (_N_=168 participants).
- Each participant played the Totem game for 25 minutes.
- Participants in Gen. 2-4 inherited **recipes** from their ancestor.

**Research questions**

1. How likely are future generations to exceed their ancestors?
2. Does inheritance have an impact on new tool discovery?

<!--
In this experiment, we recruited 168 participants and assigned them to 42
complete four-generation chains. Each participant played the Totem game for 25
minutes. Participants in Generation 2 through 4 started off with the
**recipes** for all tools discovered by their ancestor. So they could see
which tools their ancestors had discovered, and they could click on them
to see the combinations that created them.

This is important because by inheriting the recipes, and not the tools,
the first thing each participant had to do was recreate the tools discovered
by their ancestor before they could attempt any new tools.

My main research question was how learning from someone else's successes was
different than discovering the same tools individually. I asked this in two
ways.

First, I asked how likely future generations are to exceed their ancestors.
After recreating the previous tools, can they continue on and discover new
ones, or do they just become stuck because they have too many items.

Second, I asked whether inheriting has any impact on new tool discovery, so
after recreating tools discovered by someone else, is your rate of new tool
discovery affected?
-->

# Tools by generation

```{r fig3, fig.width=5, fig.height=5, out.height="90%"}
innovations_by_generation_plot
```

<!--
First I'll just show you the results by generation. So here each thin line is a
chain, and you can see that each generation tended to discover more tools in 25
minutes than their ancestor.

You can see that this is a quadratic fit, indicating that later generations had
a harder time exceeding their ancestors than earlier generations.
-->

# Tools by inheritance size

```{r fig4}
grid.arrange(
  innovations_created_and_inherited_plot,
  new_innovations_plot,
  ncol = 2
)
```

<!--
Next I'm going to show you the same data, but this time based on the number
of inherited items, rather than by generation. The reason is that there is
no difference between a second generation participant inheriting 6 tools and
a fourth generation participant inheriting the same 6 tools.

On the left you can see the relationship between number of items inherited
and number to tools discovered. The dotted line is a reference line with a
slope of 1, so people who fall along that line recreated the tools they
inherited but were unable to create any new tools. As you can see, most of
the points are above the line, but to get a better sense of the relationship,
look on the right, where the same reference line is now horizontal, and you
can see the negative slope best fit line, indicating that as participants
inherited more tools, they had a harder time exceeding those tools.
-->

# Learning times

```{r fig5, fig.width=5, fig.height=5, out.height="90%"}
learning_rates_plot
```

<!--
Now I want to dig in to what explains why inheriting more tools seemed to
decrease future problem solving ability.

The first way to do this is to take into account how much time in the 25 minute
session participants spent recreating existing tools as opposed to discovering
new ones.

Here I'm showing you the relationship between the size of the inheritance,
and the amount of time spent recreating the inherited tools, and you can see
the expected positive relationship, indicating that more tools took proportionally
longer to recreate.
-->

# Discovery rates

```{r fig6, fig.width=5, fig.height=5, out.height="90%"}
playing_time_plot
```

<!--
Now we can ask how effectively participants were able to discover new tools
taking into account how much time it took them to recreate the old tools.

Here I'm showing that people who had more time to discover new tools tended
to discover more tools, but what's interesting is that the slope of this line
does not vary based on the number of tools that were inherited. That suggests
that even participants who inherited a bunch of tools were not any slower
at discovering new tools, they just had less time to do so.
-->

# Guesses per tool

```{r fig7}
grid.arrange(
  guesses_per_item_by_inheritance_plot + ggtitle("Inherited tools"),
  guesses_per_new_item_by_inheritance_plot + ggtitle("Discovered tools"),
  ncol = 2
)
```

<!--
The last analysis I'll show breaks things down further by the number of guesses
made for each tool. In the previous analyses, I treated all tools the same,
but here I'm saying, for each tool, from the moment at which the tool was available
to be discovered, how many guesses did it take to discover it.

Here I've split things up based on whether the tool was inherited from an ancestor,
or it had to be discovered new. So on the left, you can see the benefit of inheriting.
Each line is a tool, and participants who inherited the recipes for those tools
were able to recreate those tools very quickly.

But the interesting test is on the right. Here I'm asking whether learning from
someone else's solutions gave you any sort of insight into the discovery of new
tools as opposed to discovering all tools on your own. The lack of an effect
here is interesting because it indicates that inheriting from an ancestor does
not have an impact on future problem solving.
-->

# Summary

1. How likely were participants able to exceed their ancestors?
  - Participants were able to solve more problems in 25 minutes than their ancestors.
2. Does inheritance have an impact on new tool discovery?
  - No effect on the rate of new tool discovery.
  - No effect on the number of guesses required for new tools.

<!--
To wrap things up, I found that participants were able to solve more problems in 25 minutes
than their ancestors. This type of iterated problem solving is known to be an important
factor in human evolution, and here we've added to our understanding of how problem solving
knowledge accumulates over generations.

To me the more interesting question is whether we can detect an impact of inheritance on
new tool discovery. However, we didn't find any effects on the rate of new tool discovery,
and neither did we find an effect on the number of guesses required for new tools.
-->

# Why isn't this the complete story?

How does iterated problem solving compare to alternative strategies?

- Is it better to pass on to the next generation or continue on individually?
- Is it better to work in sequence or in parallel?

```{r team-structures, out.height="60%", fig.align="center"}
crotchet::draw_graphviz("team-structures", package = "totems")
```

<!--
This work is just starting, I want to briefly mention some of the other
questions we have using this paradigm. The main limitation to me is that we
might get more understanding about iterated problem solving if we compare an
iterated strategy to alternative ways of solving the same set of problems. That
allows us to answer questions like: Is it better to pass on to the next
generation or continue on individually? and Is it better to work in sequence or
in parallel?

I have some of these results, but didn't have time to talk about them here,
so if you are interested in them, please feel free to ask me about them after.
-->

# Cumulative improvements in iterated problem solving

**Pierce Edmiston**, Max Derex, Gary Lupyan  
pedmiston@wisc.edu  
osf.io/vf2wk  
[github.com/pedmiston/iterated-problem-solving](https://github.com/pedmiston/iterated-problem-solving)

# References {.allowframebreaks}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.25in}
\noindent
\footnotesize
